---
title: "Class 18: Mini Project"
author: "Rachel Lamm (A:18518313)"
format: pdf
toc: true
---

## Investigating pertussis cases by year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(datapasta)
library(ggplot2)
cdc<- data.frame(
                                          Year = c(1922L,1923L,1924L,1925L,1926L,
                                                   1927L,1928L,1929L,1930L,
                                                   1931L,1932L,1933L,1934L,
                                                   1935L,1936L,1937L,1938L,
                                                   1939L,1940L,1941L,1942L,
                                                   1943L,1944L,1945L,1946L,
                                                   1947L,1948L,1949L,1950L,
                                                   1951L,1952L,1953L,1954L,1955L,
                                                   1956L,1957L,1958L,1959L,
                                                   1960L,1961L,1962L,1963L,
                                                   1964L,1965L,1966L,1967L,
                                                   1968L,1969L,1970L,1971L,
                                                   1972L,1973L,1974L,1975L,
                                                   1976L,1977L,1978L,1979L,
                                                   1980L,1981L,1982L,1983L,1984L,
                                                   1985L,1986L,1987L,1988L,
                                                   1989L,1990L,1991L,1992L,
                                                   1993L,1994L,1995L,1996L,
                                                   1997L,1998L,1999L,2000L,
                                                   2001L,2002L,2003L,2004L,
                                                   2005L,2006L,2007L,2008L,
                                                   2009L,2010L,2011L,2012L,2013L,
                                                   2014L,2015L,2016L,2017L,
                                                   2018L,2019L,2020L,2021L,
                                                   2022L,2023L),
                  No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                                   202210,181411,161799,197371,
                                                   166914,172559,215343,
                                                   179135,265269,180518,147237,
                                                   214652,227319,103188,
                                                   183866,222202,191383,191890,
                                                   109873,133792,109860,156517,
                                                   74715,69479,120718,68687,
                                                   45030,37129,60886,62786,
                                                   31732,28295,32148,40005,
                                                   14809,11468,17749,17135,
                                                   13005,6799,7717,9718,4810,
                                                   3285,4249,3036,3287,1759,
                                                   2402,1738,1010,2177,2063,
                                                   1623,1730,1248,1895,
                                                   2463,2276,3589,4195,2823,
                                                   3450,4157,4570,2719,4083,
                                                   6586,4617,5137,7796,6564,
                                                   7405,7298,7867,7580,9771,
                                                   11647,25827,25616,15632,
                                                   10454,13278,16858,27550,
                                                   18719,48277,28639,32971,
                                                   20762,17972,18975,15609,
                                                   18617,6124,2116,3044,7063))
              
```

```{r}
ggplot(cdc) +
  aes(x=Year, y= No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(
    title = "Pertussis Cases by Year (1922-2019)",
    x = "Year",
    y = "Number of Cases"
     )
```
## A tale of two vaccines (wP & aP)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(x=Year, y= No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, linetype = "dashed", color ="blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", color ="red") +
  geom_text(aes(x = 1946, y = max(No..Reported.Pertussis.Cases), label = "wP"), angle = 90, vjust = -0.5, color ="blue") +
  geom_text(aes(x = 1996, y = max(No..Reported.Pertussis.Cases), label = "aP"), angle = 90, vjust = -0.5,color ="red") +
  labs(
    title = "Pertussis Cases by Year (1922-2019)",
    x = "Year",
    y = "Number of Cases"
     )
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, Pertussis cases began to increase again. This may be be due to the warning immunity from the aP vaccine, bacterial evolution, vaccine hesitancy, or more senstive testing. 

## Exploring CMI-PB data

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```

```{r}
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate) 
```

```{r}
subject$year_of_birth <- ymd(subject$year_of_birth)
subject$age <- today() - subject$year_of_birth
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")
round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length(wp$age, "years") ) )
```

```{r}
t.test(time_length(ap$age, "years"), time_length(wp$age, "years"))
```

Yes, they are significantly different.

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

Yes, they are significantly different.


```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer    <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE)
```



>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- left_join(specimen, subject)
dim(meta)
head(meta)
```


>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta, by = "specimen_id")
dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The adbata table has 4 different data sets from 2020-2023. The number of rows decreases as the years increase seeing as 2023 has the fewest entries. 

## Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(x= MFI, y= antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others? 

From this ggplot is looks like PT, PRN, FHA, and TT show differences over time. I think these show more than others because the others aren't big factors or components of the vaccines.  

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(x = MFI, col = infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen == "FIM2/3") %>%
  ggplot() +
  aes(x = MFI_normalised, col = infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

>Q16. What do you notice about these two antigens time courses and the PT data in particular?

I notice that PT IgG titers rise over time and peak around 5-6. They are much higher than the control antigen OVA which shows a strong vaccine induced response. FIM2/3 titers are low with not much change. 

>Q17. Do you see any clear difference in aP vs. wP responses?

No, not any significant difference. 

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

>Q18. Does this trend look similar for the 2020 dataset?

Yes, this trend looks similar. 

## Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

>Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(x = visit, y = tpm, group = subject_id) +
  geom_point() +
  geom_line(alpha = 0.2)
```

>Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

I notice that it peaks shortly after the booster which seems to be around visit 4.

>Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

Yes, it does match the trend because they both peak at around 4/5.

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```

