---
title: "Class 11: AlphaFold"
author: "Rachel Lamm (A18518313)"
format: pdf
toc: true
---

## Background

We saw last day that the main respitory for biomolecular structure (the PDB database) only has ~250,000 entries 

UniProtKB (the main protein sequence database) has over 200 million entries!


## Alphafold 

In this hands-on session we will utilize AlphaFold to predict protein structure from sequence (Jumper et al. 2021).

Without the aid of such approaches, it can take years of expensive laboratory work to determine the structure of just one protein. With AlphaFold we can now accurately compute a typical protein structure in as little as ten minutes.




## The EBI AlphaFold databse

The EBI alphafold database contains lots of computed structure models increasingly likely that the structure you are interested in is already in this database at < https://alphafold.ebi.ac.uk

There are 3 major outputs from AlphaFold

1. A model of structure in **PDB** format
2. a **pLDDT score**: that tells us how confident the model is for a given residue in your protein (High values are good, above 70)
3. a **PAE score** that tells us about protein packing quality 

If you can't find a matching entry for the sequence you are interested in AFDB you can run AlphaFold yourself...


## Running AlphaFold

We will use ColabFold to run AlphaFold on our sequence

Figure from AlphaFold here!

## Interpreting results

Custom analysis of resulting models

We can read all the AlphaFold results into R and do more quantitative analysis than just viewing the structures in Mol-star:

Read all the PDB MODELS:

```{r}
class11_dir <- "hipvrdimer_94b5b/"
```

```{r}
pdb_files <- list.files(path=class11_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

basename(pdb_files)
```

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```
```{r}
pdbs
```
```{r}
rd <- rmsd(pdbs, fit=T)
```

```{r}
range(rd)
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```


## Predicted Alignment Error for domains


```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=class11_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```
```{r}
head(pae1$plddt) 
```

```{r}
pae1$max_pae
pae5$max_pae
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```
```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

## Residue conservation from alignment file

```{r}
aln_file <- list.files(path=class11_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

How many sequences are in this alignment

```{r}
dim(aln$ali)
```

```{r}
sim <- conserv(aln)
```

```{r}
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```







